---
title: "LAI within Riparian Restorations, Final Project Data Sci Fall 2024"
output: pdf_document
date: "2024-11-22"
author: Holly Amer
---

```{r setup, include=FALSE}
 knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# all necessary packages
library(ggplot2)
library(ggpubr)
library(tidyverse)
library(broom)
library(lme4)
library(ggpmisc)
library(sf)
library(viridis)
library(ggspatial)
library(car)
library(lme4)
library(Matrix)
library(performance)

```


# Introduction
1)  Includes relevant background and references to scientific literature that foreshawdow the importance of each question;
 
 There have been more than 2,000 riparian restoration projects that have been planted over the last 30 years across Oregon. These projects are costly when it comes to money and time, so we want them to be well planned and reach the highest potential possible. They are beneficial to ecosystems because they positively impact water, wildlife, and humans that live near them. By researching riparian restorations across Oregon and looking at their growth over time, we can see which soil texture and water availability strata have the best effect on the plants within them. This will allow us to see which set of conditions will have the highest plant growth over time. 
 
  From my current research, there is data from riparian restorations within the western side of Oregon including the ecoregions of the Willamette Valley, Coast Range, and Klamath Mountains. The data collected ranges across different soil textures (fine/coarse), water stress (low/high), and age groups (0-5, 5-10, 10-15, 15-20, and 20+). It aims to study how LAI will differ through time, across Oregon, through soil textures and water availability. This research will tell us the most effective spatial areas in Oregon to conduct efforts in riparian restoration to draw down as much carbon as possible, and create essential refined spatial data to inform future climate mitigation decisions.
 
  There are many studies on Riparian Restorations, Western Oregon, and LAI as separate entities in many formats, but none that encompass all three areas. Detailed empirical data relating LAI and riparian restoration success has not been collected before in western Oregon. Soil texture and water stress are two variables that are known to affect plant growth and should be measured to assess their impacts on restoration projects within Oregon. To understand restoration trajectories, Leaf Area Index (LAI), a method that is widely used for measuring percent canopy cover, can be used to assess riparian planting success. Taking Digital Hemispherical Photos (DHP) allows researchers to have a permanent trace of data, which can be used for future comparisons and analyses (Fang et al., 2019, Chianucci and Cutini, 2012). LAI can be compared to variables such as soil texture and water stress, to assess how they impact plant growth over time. Restoration has been focused on as a primary goal in climate change mitigation for its positive impacts on biodiversity, wildlife habitats, and stream health. Carbon sequestration is a strong co-benefit of restoration that should be maximized (Dybala and Steger et al. 2018). 
  
  It is essential that ecologists measure the success of current riparian restoration projects, as we move into a future where restoration will continue to be a necessity for the health of our environment. So far, riparian restoration research has been spatially focused on North America, with South America with the next most studies (Dybala, Matzek, et al., 2018). There is also more focus on smaller temporal scales and fewer long-term studies that track ecosystem responses through climate change impacts (Li et al., 2022). Although more difficult, having studies over longer periods will help us to see how climate change affects these ecosystems, and which variables are the most affected. 
  


2)  a clear statement of 2-3 hypotheses. Using the OBA data, the hypotheses should address understanding a new dimension of bee conservation, ecology, or management. The hypotheses should be logical (i.e., make sense given published literature). 

 My questions include: What is the impact of soil texture, water stress, and age on Leaf Area Index (LAI) in Riparian Restoration projects in Western Oregon?

  My hypotheses are that as plantings become older, soil texture becomes more fine, and water stress is lower, LAI will increase. 


# Data
1) List of the three+ different data sets used, the type of data each represented, and general metadata (same as in the project workflow).
  I will primarily be using my own data. This includes Site info like water stress conditions, soil texture percentages, and age group, and the LAI values within each site/plot. 

2). The project should then read in and tidy the 2028-2023 data (including cleaning plant taxonomy if plant data is used, etc.) 

# Combining and Cleaning the data
```{r load in data}
# load in LAI data
LAI <- read.csv("RR_TNC_LAI_MASTER.csv", header = TRUE, colClasses = c("factor", "factor", "factor", "factor", "factor", "factor", "numeric", "factor", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric"))

#texture 2023
texture2023 <- read.csv("texture_2023.csv", header = TRUE)

#texture 2024
texture2024 <- read.csv("texture_2024.csv", header = TRUE)

# r data file with water stress contents
load("text_water.RData")


```

```{r cleaning the texture dataset}
# cleaning the texture dataset

# selecting relevant columns
texture2023 <- texture2023 %>% select(site, plot, sand, clay, silt)

# selecting relevant columns
texture2024 <- texture2024 %>% select(site, plot, sand, clay, silt)

# combining 2023 and 2024 datasets
texture <- rbind(texture2023, texture2024)

# group texture by site and take averages
texture <- texture %>%
  group_by(site) %>%
  summarize(
    avg_silt = mean(silt, na.rm = TRUE),
    avg_sand = mean(sand, na.rm = TRUE),
    avg_clay = mean(clay, na.rm = TRUE)
  )


# drop sites with poor data/no photos
texture <- texture %>%
  filter(!(site %in% c("DAMa", "DAMb", "FTE", "HC20a", "HF10b", "LC00a", "LC10b", "LF05e", "LF15d", "LF15e", "PRV", "WC")))

```

```{r cleaning the water stress dataset}
# cleaning the waterstress dataset

# group water stress by site and take averages
climatedeficit <- climatedeficit %>%
  group_by(site) %>%
  summarize(
    avg_deficit = mean(deficit, na.rm = TRUE))

# drop sites with poor data/no photos
climatedeficit <- climatedeficit %>%
  filter(!(site %in% c("HC00a", "HC00b", "HF00b", "HF15a", "LF00b", "HF00c", "DAMa", "DAMb", "FTE", "HC20a", "HF10b", "LC00a", "LC10b", "LF05e", "LF15d", "LF15e", "PRV", "WC", "LF20d")))

# check same number of sites in each dataset
nrow(climatedeficit) == nrow(texture)

```



# Data Wrangling and Summarizing
Manipulation and summarizing data in relevant ways using tables, scatter plots,  histograms, barplots, etc. Initial exploration of data.

```{r soil texture triangle and sites}
# soil texture triangle showing sites by texture type and age

# LAI and texture
texture_LAI <- merge(texture, LAI, by = "site")

master_dataset <- merge(texture_LAI, climatedeficit, by = "site")

#creating texture triangle
library(ggtern)
library(ggplot2)
theme_set(theme_bw())
master_dataset %>%
  ggtern(aes(
    x = avg_sand,
    y = avg_clay,
    z = avg_silt,
    #color = years
    color = avg_deficit
  )) + 
  geom_point(size = 5) + 
  theme_showarrows() +
    labs(yarrow = "clay (%)",
       zarrow = "silt (%)",
       xarrow = "sand(%)",
       title = "Sites by Soil Texture and Water Deficit") +
  scale_colour_gradient(low = "#3288bd",
                        high = "#d53e4f", 
                        name = "Average Water Deficit") +
  theme_clockwise() +
  theme_hidetitles() +
  theme( plot.title = element_text(size = 16, face = "bold"), 
    legend.title = element_text(size = 14, face = "bold"), 
    legend.text = element_text(size = 12))

# This code was taken from the following source:
# https://saryace.github.io/flipbook_soiltexture_en/#1

```
```{r summary of water stress by age}
# grouping water stress
# deficit = PVT - EVT
# < 225 mm lower water stress, >225 mm higher water stress

master_dataset <- master_dataset %>%
  mutate(
    water_stress_group = case_when(
      avg_deficit <= 225 ~ "Low",
      avg_deficit > 225 ~ "High"
    )
  )

# summarizing
climate_summary <- master_dataset %>%
  group_by(water_stress_group) %>%
  summarize(num_sites = n_distinct(site))



# Bar graph for climate type
ggplot(climate_summary, aes(x = water_stress_group, y = num_sites, fill = water_stress_group)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Number of Sites by Climate Type",
    x = "Climate Type",
    y = "Number of Sites"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r scatterplot of lai values based on water stress and texture}
# average LAI by site 
master_dataset <- master_dataset %>%
  group_by(site) %>%
  mutate(LAI_4_ring_avg = mean(LAI_4_ring, na.rm = TRUE)) %>%
  ungroup()

# scatterplot of lai values based on water stress and texture

library(viridis)

ggplot(master_dataset, aes(
  x = years, 
  y = LAI_4_ring_avg, 
  color = avg_sand, 
  shape = water_stress_group
)) +
  geom_point(size = 3) + 
  scale_color_viridis() +
  labs(
    x = "Years", 
    y = "Average LAI", 
    color = "Average Sand (%)", 
    shape = "Water Stress Group",
    title = "Average LAI at Each Site by Sand %, Age, and Water Stress"
  ) +
  theme_minimal()
  




```


# Hypothesis test
Use of a statistical test (simulation of the null hypothesis, A/B testing or any other appropriate test) to test the stated hypotheses. 

Correct interpretation of the outcome of the test. 
```{r}
# This chunk is me testing different models, and looking at model improvement as I go. This is a WIP.

#add column for silt+clay texture
master_dataset <- master_dataset %>% mutate(soiltexture = avg_silt+avg_clay)

#scale and center all numeric data
lai_scale <- master_dataset%>% mutate_if(is.numeric, scale)


# model with just center/streambank
model1 <- lmer(LAI_4_ring ~ center_streambank + (1 | site/plot), data = lai_scale)
summary(model1)

# add planting age
model2 <- lmer(LAI_4_ring ~ center_streambank * years + (1 | site/plot), data = lai_scale)
summary(model2)

# compare models 1 and 2
anova(model1, model2) #improvement

# add soil texture
model3 <- lmer(LAI_4_ring ~ center_streambank * years * soiltexture + (1 | site/plot), data = lai_scale)
summary(model3)

# compare models 2 and 3
anova(model2, model3) #improvement, p-value highly significant <.001 
# This model tells us that water stress is related to soil texture depending on time and photo position within plot (center or streambank)

# add water stress 
# model4 <- lmer(LAI_4_ring ~ center_streambank * years * soiltexture + avg_deficit + (1 | site/plot), data = lai_scale)
# summary(model3) 
# anova(model3, model4) # does not improve model


# plot predicted vs actual LAI values
# calculate predicted LAI with scaled data
lai_scale$predicted_LAI <- predict(model3)

# back-transform the predicted LAI to the original scale
original_mean <- mean(master_dataset$LAI_4_ring, na.rm = TRUE)
original_sd <- sd(master_dataset$LAI_4_ring, na.rm = TRUE)

lai_scale$predicted_LAI_original <- (lai_scale$predicted_LAI * original_sd) + original_mean

# add observed (collected) LAI from the unscaled dataset
lai_scale$collected_LAI <- master_dataset$LAI_4_ring

# create the new dataset with the required columns
new_dataset <- lai_scale %>%
  select(site, plot, predicted_LAI_original, collected_LAI)# get predicted values using the scaled dataset (lai_scale)

master_dataset$predicted_LAI <- predict(model3)



# plot predicted vs observed
ggplot(new_dataset, aes(x = predicted_LAI_original, y = collected_LAI)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "red") +
  theme_minimal() +
  xlab("Predicted LAI") +
  ylab("Observed LAI") +
  ggtitle("Predicted vs. Observed LAI") +
  
   stat_poly_eq(
    aes(label = paste(..eq.label.., ..rr.label.., ..p.value.label.., sep = "~~~")), # add p-value and r values to graph
    formula = y ~ x,
    parse = TRUE,
    label.x = 5, 
    label.y = 5 
  )
#model fits well



```



# Visualization
Final and publication-worthy visuals: They should address the hypotheses test, be clear, have correct labels, and include legends and units. 
```{r map of LAI points}

# keep relevant data for mapping
LAI_mapping <- master_dataset %>% select(site, plot, avg_sand, years, latitude, longitude, water_stress_group)

# create a spatial object
LAI_sf <- st_as_sf(LAI_mapping, coords = c("longitude", "latitude"), crs = 4326)


# add ecoregions to map
ecoregions <- st_read("OR-ecoregions/Ecoregions_OregonConservationStrategy.shp")

# only relevant ecoregions
ecoregions_filtered <- ecoregions %>%
  filter(Ecoregion %in% c("Willamette Valley", "Klamath Mountains", "Coast Range"))

# make map
ggplot() +
  
  # ecoregions 
  geom_sf(data = ecoregions_filtered, aes(fill = Ecoregion), color = "black", alpha = 0.2) +
  
  # LAI points
  geom_sf(data = LAI_sf, aes(color = avg_sand, shape = water_stress_group), size = 3) +
  scale_color_viridis() + 
  labs(color = "Average Sand (%)", shape = "Water Stress Group") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_x_continuous(breaks = seq(-125, -120, by = 1)) +
  ggtitle("Map of LAI points by Ecoregion with Texture \n and Water Stress") +


# add a north arrow 
  annotation_north_arrow(location = "tl", which_north = "true", height = unit(.8, "cm"), width = unit(.7, "cm")) +
    
# add a scale
  annotation_scale(location = "bl", width_hint = 0.3, height = unit(.1, "cm"))





```
```{r}
# LAI vs texture simple model
model_texture <- lm(LAI_4_ring_avg ~ soiltexture, data = master_dataset)
summary(model_texture)



# LAI vs texture plot
ggplot(master_dataset, aes(x = soiltexture, y = LAI_4_ring_avg)) +
  geom_point(color = "blue", alpha = 0.7, size = 3) + 
  geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "dashed") + # add regression line
  
  stat_poly_eq(
    aes(label = paste(..eq.label.., ..rr.label.., ..p.value.label.., sep = "~~~")), # add p-value and r values to graph
    formula = y ~ x,
    parse = TRUE,
    label.x = 1, 
    label.y = 3 
  ) +

  labs(
    title = "Relationship Between Soil Texture and LAI",
    x = "Soil Texture (Clay+Silt%)",
    y = "Leaf Area Index (LAI)"
  ) +
  theme_minimal() +
  theme( plot.title = element_text(size = 16, face = "bold"), 
    legend.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 14), 
    axis.title = element_text(size = 16))



# model water stress simple 
model_waterdeficit <- lm(LAI_4_ring_avg ~ avg_deficit, data = master_dataset)
summary(model_waterdeficit)

# LAI vs water stress
ggplot(master_dataset, aes(x = avg_deficit, y = LAI_4_ring_avg)) +
  geom_point(color = "blue", alpha = 0.7, size = 3) + 
  geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "dashed") + # add regression line
  
    
  stat_poly_eq(
    aes(label = paste(..eq.label.., ..rr.label.., ..p.value.label.., sep = "~~~")), # add p-value and r values to graph
    formula = y ~ x,
    parse = TRUE,
    label.x = 5, 
    label.y = 5 
  ) +
  
  labs(
    title = "Relationship Between Water Stress and LAI",
    x = "Water stress (PVT-EVT)",
    y = "Leaf Area Index (LAI)"
  ) +
  theme_minimal() +
  theme( plot.title = element_text(size = 16, face = "bold"), 
    legend.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 14), 
    axis.title = element_text(size = 16))

# LAI vs age plot
ggplot(master_dataset, aes(x = years, y = LAI_4_ring_avg)) +
  geom_point(color = "blue", alpha = 0.7, size = 3) + 
  geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "dashed") + # add regression line
  
  stat_poly_eq(
    aes(label = paste(..eq.label.., ..rr.label.., ..p.value.label.., sep = "~~~")), # add p-value and r values to graph
    formula = y ~ x,
    parse = TRUE,
    label.x = 1, 
    label.y = 3 
  ) +

  labs(
    title = "Relationship Between Age and LAI",
    x = "Site Age",
    y = "Leaf Area Index (LAI)"
  ) +
  theme_minimal() +
  theme( plot.title = element_text(size = 16, face = "bold"), 
    legend.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 14), 
    axis.title = element_text(size = 16))

```


# Conclusion
Conclusion that clearly answers the original hypotheses.
Leaf Area Index can be influenced by many different variables, as canopy cover and plant life thrive under different conditions, depending on their preferred habitats. Some of the original data exploration shows where the variation in my data lies, including more low water stress than high water stress, and many silty sites sampled. Through the data analysis I have completed, I show that there was a strong significance of site age on LAI, with a p-value of <.001 and an R^2 of .47, which aligned with my hypothesis. I also showed that there is not a significant influence between LAI and soil texture, or LAI and water stress alone. I had originally hypothesized that sites with finer soil would have a higher LAI, and sites with lower water stress would also have a higher LAI value, but these trends were not shown in individual linear regression models. However through a mixed linear effects model, LAI is related to soil texture, with the slope of that relationship varying depending on the center/streambank position, and the planting age. This is shown through the results of model 3, which has a strong model significance with an R^2 of .77 and a p-value of <.001. This means that the chance of seeing these results based on chance is less than .1%, and we can reject the null hypothesis. Our high R^2 value tells us that 77% of the variation in LAI can be explained by soil texture, age, and image placement together. I plan on doing further analysis to this data, such as testing variables like temperature and precipitation to see if they have an impact on LAI rather than the water stress variable, and playing around with other current model interactions. 

# References
Acevedo, Sara. Soil texture triangles using R. (2021). 
https://saryace.github.io/flipbook_soiltexture_en/#1

Chianucci, F., & Cutini, A. (2012). Digital hemispherical photography for estimating forest canopy properties: Current controversies and opportunities. Iforest-Biogeosciences and Forestry, 5, 290–295. https://doi.org/10.3832/ifor0775-005 

Fang, H., Baret, F., Plummer, S., & Schaepman-Strub, G. (2019). An overview of global leaf area index (LAI): Methods, products, validation, and applications. Reviews of Geophysics. 57, 739–799. https://doi.org/10.1029/2018RG000608 

Dybala KE, Matzek V, Gardali T, Seavy NE. Carbon sequestration in riparian forests: A global synthesis and meta-analysis. Glob Change Biol. 2018; 25: 57–67. https://doi.org/10.1111/gcb.14475 

Dybala KE, Steger K, Walsh RG, Smart DR, Gardali T, Seavy NE. Optimizing carbon storage and biodiversity co-benefits in reforested riparian zones. J Appl Ecol. 2019; 56: 343–353. https://doi.org/10.1111/1365-2664.13272 

Li, W., Migliavacca, M., Forkel, M. et al. Widespread increasing vegetation sensitivity to soil moisture. Nat Commun 13, 3959 (2022). https://doi.org/10.1038/s41467-022-31667-9 

Thorson, T.D., Bryce, S.A., Lammers, D.A., Woods, A.J., Omernik, J.M., Kagan, J., Pater, D.E., and Comstock, J.A., 2003. Ecoregions of Oregon (color poster with map, descriptive text, summary tables, and photographs): Reston, Virginia, U.S. Geological Survey (map scale 1:1,500,000).



